Ubiquitous Digital Technologies and Spatial Structure; an update

### 2020-12-01 paper/2_resubmission/Tranos Ioannides Revision November 19 2020.tex
responses

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the Zipf
coefficient for a broad sample of countries over time. Zipf coefficient
is one of the most widely used measures of spatial dispersion with
numerous applications in urban economics and economic geography ({[}see
for example {[}36{]}; {[}37{]}; {[}38{]}; {[}39{]}; {[}40{]}; {[}41{]};
{[}42{]}). This an appropriate measure of dispersion because of the
extreme heterogeneity of the city size distribution and the very good
fits normally obtained with such estimations. Skipping details
pertaining of the suitability of the approach, which may be found in
Ioannides et al.~{[}18{]} and Tranos and Ioannides {[}19{]}, we proceed
with the estimation of the logarithmic form of:

@@

This leads to an estimate of \(\zeta\), known as the Zipf coefficient,
but more appropriately the Pareto exponent, when \(\zeta\) is not
estimated to be near \(1\). While much of the literature focuses on the
Zipf coefficient being close to \(1\), several estimations of city size
distributions lead to estimates of \(\zeta\) that are not necessarily
equal to \(1\), or even near it, in which case we refer to \(\zeta\) as
the Pareto, or power law, exponent. Given that our aim here is to
estimate \(\zeta\) for a number of countries over time as a measure of
dispersion, equation \ref{eq1} describes the rank of city \(i\) in
country \(c\) in year \(t\):

@@

\caption{\label{tab:unnamed-chunk-2}Zipf coefficients\label{zipf}}

@@

Zipf coef. & 2000 & 2001 & 2002 & 2003 & 2004 & 2005 & 2006 & 2007 & 2008 & 2009 & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018\\

@@

\begin{table}[!htbp] \centering 
  \caption{Descriptive statistics for the multi-country model\label{desc.global}} 
  \label{} 
\footnotesize 
\begin{tabular}{@{\extracolsep{1pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Zipf coefficient & 950 & $-$1.9 & $-$1.1 & 0.2 & $-$0.7 \\ 
Zipf inv. sq. error\textsuperscript{a} & 950 & 0.01 & 0.1 & 0.05 & 0.2 \\ 
Population density & 931 & 2.5 & 126.2 & 182.1 & 1,239.6 \\ 
Government expenditure (\% GDP) & 913 & 1.0 & 14.7 & 4.9 & 30.0 \\ 
Trade (\% GDP) & 914 & 19.1 & 65.3 & 33.9 & 220.4 \\ 
GDP growth & 934 & $-$33.1 & 4.0 & 4.2 & 54.2 \\ 
GDP per capita (log) & 892 & 630.7 & 18,049.2 & 15,432.9 & 61,391.4 \\ 
Female labour force (\%) & 950 & 7.9 & 37.8 & 11.9 & 55.2 \\ 
Population & 950 & 5,122,493 & 116,073,037.0 & 244,323,372.0 & 1,392,730,000 \\ 
Non agriculture value added (\% GDP) & 930 & 58.8 & 90.3 & 8.2 & 99.4 \\ 
Internet users per 100 hab. & 916 & 0.01 & 32.5 & 28.2 & 96.0 \\ 
Broadband users per 100 hab. & 824 & 0.0 & 8.8 & 11.2 & 44.8 \\ 
Mobile phone users per 100 hab. & 949 & 0.0 & 72.9 & 46.4 & 191.0 \\ 
Fixed phone users per 100 hab. & 946 & 0.0 & 19.8 & 18.6 & 68.4 \\ 
\hline \\[-1.8ex] 
\multicolumn{6}{l}{\textsuperscript{a}This is the inverse squared standard error of the estimated Zipf, coefficient, which is used for weighting} \\ 
\multicolumn{6}{l}{the observations for the estimation, of equation 3. See Section 4} \\ 
\end{tabular} 
\end{table}

@@

The availability of a panel dataset for city sizes across countries
enables us to use country fixed effects, which can address potential
endogeneity issues related to unobserved country specific
characteristics of city size distributions. However, such a strategy
does not address potential simultaneity issues. Simply put, internet
penetration might be affected by spatial structure, as reflected in Zipf
coefficients, or both internet penetration and spatial structure might
be jointly determined by a third variable. E.g., if a country already
has a dispersed spatial structure, internet is particularly suitable in
facilitating communication. Potential endogeneity in our specification
will prevent us from being able to determine the causal impact of
internet and digital communication technologies usage on spatial
structure, which is the main aim of this paper. In order to address this
problem, we will adopt an instrumented variable strategy. Table
\ref{desc.global} also includes the descriptive statistics for the
instrumental variable we are using and (\(female\:labour\:force\)), will
be discussed in \protect\hyperlink{sec4}{Section 4}.

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the Zipf coefficient, as estimated according to
the Gabaix and Ibragimov {[}44{]} correction. The main explanatory
variables of interest are provided at the top of Table \ref{ols.global};
namely internet, broadband, mobile and fixed telephony per \(100\)
habitants, expressed in natural logarithms. The main variables of
interest are introduced successively on their own in the regressions
reported in Table \ref{ols.global}. All regressions include country
fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared
standard error of the estimated Zipf coefficient (see Table
\ref{desc.global}) to address potential noise that is carried over from
the first part of our identification strategy. In regards to the
interpretation of the estimated coefficients, given that the Zipf
coefficient has entered the regression not as an absolute value, but
instead as a real number a positive coefficient for a RHS variable
indicates an impact towards the decrease of the spatial dispersion of
population. In other words, a positive coefficient indicates an effect
towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies
because of the expansion of digital technologies.

\begin{table}[!htbp] \centering 
  \caption{OLS estimation of equation (3)\label{ols.global}} 
  \label{} 
\small 
\begin{tabular}{@{\extracolsep{1pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Zipf Coefficient 2000-18} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Internet users per 100 hab. (log) & 0.001$^{***}$ &  &  &  \\ 
  & (0.0001) &  &  &  \\ 
  Broadband users per 100 hab. (log) &  & $-$0.001$^{**}$ &  &  \\ 
  &  & (0.0002) &  &  \\ 
  Mobile phone users per 100 hab. (log) &  &  & 0.0003 &  \\ 
  &  &  & (0.0003) &  \\ 
  Fixed phone users per 100 hab. (log) &  &  &  & $-$0.00003 \\ 
  &  &  &  & (0.0001) \\ 
  Population density (log) & $-$1.378 & $-$1.060 & $-$1.505 & $-$1.892 \\ 
  & (1.870) & (1.985) & (1.896) & (2.007) \\ 
  Government expenditure (% GDP) & 0.0002 & 0.003$^{***}$ & 0.001 & 0.001 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  Trade (\% of GDP) & 0.0002$^{***}$ & $-$0.0001 & 0.0001 & 0.0001 \\ 
  & (0.0001) & (0.0001) & (0.0001) & (0.0001) \\ 
  Non agriculture value added (\% GDP) & 0.0001 & 0.001 & $-$0.0001 & $-$0.0002 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  GDP growth & 0.001$^{***}$ & 0.001$^{***}$ & 0.001$^{**}$ & 0.001$^{**}$ \\ 
  & (0.0003) & (0.0004) & (0.0004) & (0.0004) \\ 
  GDP per capita (log) & $-$0.020$^{***}$ & $-$0.023$^{***}$ & $-$0.019$^{***}$ & $-$0.016$^{**}$ \\ 
  & (0.007) & (0.006) & (0.007) & (0.008) \\ 
  Population (log) & 1.300 & 0.955 & 1.390 & 1.785 \\ 
  & (1.873) & (1.985) & (1.899) & (2.011) \\ 
  Constant & $-$18.795 & $-$13.969 & $-$19.948 & $-$25.516 \\ 
  & (26.314) & (27.886) & (26.670) & (28.261) \\ 
 \hline \\[-1.8ex] 
Country fixed effects & Yes & Yes & Yes & Yes \\ 
Yearly fixed effects & Yes & Yes & Yes & Yes \\ 
Observations & 844 & 757 & 865 & 867 \\ 
Adjusted R$^{2}$ & 0.987 & 0.988 & 0.986 & 0.986 \\ 
Residual Std. Error & 0.516 & 0.497 & 0.540 & 0.540 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{4}{r}{Robust Std. Errors in parenthesis} \\ 
\end{tabular} 
\end{table}

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the Zipf coefficient, might be affected by another source,
which also affects internet penetration. For instance, economic
development might affect the concentration of population in large cities
and at the same time enable more people to go online. If we do not
address this issue, the coefficient for the main variable of interest
will capture potential effects that internet penetration has on spatial
structure, but also potential reverse causality effects that spatial
structure might generate on internet penetration. To overcome this
potential problem, Table \ref{2sls.global} reports estimates of equation
\ref{eq3} using two-stage least squares (2SLS) with instrumental
variables (IVs). The latter are variables which are correlated with our
endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any
-- of the internet and digital communications on spatial structure. At a
first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs
and the other control variables are used instead of the endogenous
variable to estimate eq. \ref{eq3}. A significant effect will verify the
causal impact of the internet and digital communication usage on spatial
structure.

\begin{table}[!htbp] \centering 
  \caption{2SLS estimation of equation (3)\label{2sls.global}} 
  \label{} 
\small 
\begin{tabular}{@{\extracolsep{1pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Zipf Coefficient 2000-18} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Internet users per 100 hab. (log) & 0.001$^{***}$ &  &  &  \\ 
  & (0.0004) &  &  &  \\ 
  Broadband users per 100 hab. (log) &  & 0.002$^{**}$ &  &  \\ 
  &  & (0.001) &  &  \\ 
  Mobile phone users per 100 hab. (log) &  &  & $-$0.007$^{*}$ &  \\ 
  &  &  & (0.004) &  \\ 
  Fixed phone users per 100 hab. (log) &  &  &  & $-$0.001$^{***}$ \\ 
  &  &  &  & (0.0003) \\ 
  Population density (log) & $-$0.827 & $-$1.736 & $-$12.765 & $-$0.828 \\ 
  & (1.841) & (2.622) & (8.326) & (2.262) \\ 
  Government expenditure (\% GDP) & $-$0.0003 & 0.0002 & $-$0.002 & $-$0.0003 \\ 
  & (0.001) & (0.002) & (0.002) & (0.001) \\ 
  Trade (\% of GDP) & 0.0004$^{***}$ & 0.0004$^{*}$ & 0.001 & $-$0.0003 \\ 
  & (0.0001) & (0.0002) & (0.0005) & (0.0002) \\ 
  Non agriculture value added (\% GDP) & 0.0003 & $-$0.0003 & $-$0.001 & $-$0.001 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  GDP growth & 0.001$^{***}$ & 0.002$^{***}$ & 0.001$^{*}$ & 0.0001 \\ 
  & (0.0004) & (0.001) & (0.001) & (0.0004) \\ 
  GDP per capita (log) & $-$0.023$^{***}$ & 0.003 & 0.034 & $-$0.0003 \\ 
  & (0.007) & (0.012) & (0.038) & (0.011) \\ 
  Population (log) & 0.776 & 1.867 & 12.794 & 0.740 \\ 
  & (1.841) & (2.646) & (8.410) & (2.267) \\ 
  Constant & $-$11.513 & $-$27.468 & $-$180.823 & $-$10.915 \\ 
  & (25.852) & (37.246) & (118.577) & (31.852) \\ 
 \hline \\[-1.8ex] 
Weak instruments & 44.77 & 22.23 & 3.51 & 27.24 \\ 
Wu-Hausman & 3 & 15.96 & 12.26 & 12 \\ 
P-value & 0.08 & 0 & 0 & 0 \\ 
Country fixed effects & Yes & Yes & Yes & Yes \\ 
Yearly fixed effects & Yes & Yes & Yes & Yes \\ 
Observations & 844 & 757 & 865 & 867 \\ 
Adjusted R$^{2}$ & 0.986 & 0.984 & 0.944 & 0.979 \\ 
Residual Std. Error & 0.534 & 0.579 & 1.078 & 0.657 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{4}{r}{Robust Std. Errors in parenthesis} \\ 
 & \multicolumn{4}{r}{IV: Female participation in labour force} \\ 
\end{tabular} 
\end{table}

@@

In total, the results indicate that our measures of internet and
telephony penetration have further enhanced agglomeration forces, at
least for large urban agglomerations, while the opposite happens for
fixed telephony. These results are robust against endogeneity issues,
but are limited to urban agglomeration included in our data. Indeed, for
the estimation of the Zipf coefficient we only included urban
agglomeration of \(300,000\) inhabitants or more due to data
availability. Hence, the above estimations cannot verify whether such an
effect is also valid for smaller cities. Interestingly, the above
results regarding the role of fixed telephony are in agreement with
previous results from {[}18{]} {[}19{]} which used a smaller panel of
countries but without constraints regarding the city size. However, our
results do not support the findings of {[}19{]} regarding internet
usage. In order to overcome the city size limitation, the next section
presents two case studies, for which we have obtained much more granular
data and therefore we are in a position to test the effect of the
internet on the tail of the urban population distribution for these
countries.

@@

We report estimation results using international country-level data on
urban agglomerations with more than \(300,000\) inhabitants from the UN
Urban Settlements data and test the effect of ICT adoption on the Zipf,
or Pareto, coefficients for national city size distributions, as
measures of dispersion for heavy-tailed data. We report results which
are robust to endogeneity concerns. Then, in order to examine such
effects for cities that are smaller than those included in our
international data, we focus on the US and the UK urban systems.
Specifically, we test the effect of internet usage and internet speed on
the changes over time in rankings of Micropolitan and Metropolitan Areas
in the US, and of Built Up Areas in the UK. The results favor a
complementary relation between the internet and agglomeration
externalities. While the cross-country estimates indicate that increase
adoption rates of such technologies has resulted in less dispersed urban
spatial structures, the two case studies reveal that internet adoption
and internet speed improved the relative population rank of a city
within its urban system. Interestingly, the latter results indicate that
such effects might be even stronger for smaller and less dense urban
areas at least in the UK. On the contrary, our estimates regarding the
effect of fixed telephony advocate towards the reverse effect: that is
fixed telephony acting against of agglomeration forces.

@@

\leavevmode\hypertarget{ref-gabaix_rank_2011}{}%
44. Gabaix X, Ibragimov R. Rank− 1/2: A simple way to improve the OLS
estimation of tail exponents. Journal of Business \& Economic
Statistics. 2011;29: 24--39.

>>>>

The present paper does not address the issue that adoption of ICT may have different effects on urban dispersion on a national scale, that is how far are major areas, that is, cities, from one another, versus dispersion within major urban areas, that is urban sprawl. Such an inquiry would require detailed data about patterns how urban sprawl  may hamper or promote economic interaction, given advancements in information technology. We think that data from the Covid-19 pandemic would likely throw light at this issue, as it is clear that not all economic activities might continue in an unencumbered manner; see Dingel and Neiman (2020).

Jonathan Dingel and Brett Neiman, "How Many Jobs Can Be Done at Home?", Becker Friedman Institute, University of Chicago, April 2020. https://tinyurl.com/y9nbmkmz

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the Pareto exponent
coefficient for a broad sample of countries over time. \textbf{The Pareto} coefficient
is one of the most widely used measures of spatial dispersion with
numerous applications in urban economics and economic geography ({[}see
for example {[}36{]}; {[}37{]}; {[}38{]}; {[}39{]}; {[}40{]}; {[}41{]};
{[}42{]}). This an appropriate measure of dispersion because of the
extreme heterogeneity of the city size distribution and the very good
fits normally obtained with such estimations. Skipping details
pertaining of the suitability of the approach, which may be found in
Ioannides et al.~{[}18{]} and Tranos and Ioannides {[}19{]}, we proceed
with the estimation of the logarithmic form of:

@@

This leads to an estimate of \(\zeta\), known as the the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere to for the remained of the text, even  when \(\zeta\) is 
estimated to be near \(1\). The term Zipf coefficient should be reserved fo the special case of \(\zeta=1\), While much of the literature focuses on the
Pareto exponent being close to \(1\), several estimations of city size
distributions lead to estimates of \(\zeta\) that are not necessarily
equal to \(1\), or even near it, in which case we refer to \(\zeta\) as
the Pareto, or power law, exponent. Given that our aim here is to
estimate \(\zeta\) for a number of countries over time as a measure of
dispersion, equation \ref{eq1} describes the rank of city \(i\) in
country \(c\) in year \(t\):

@@

\caption{\label{tab:unnamed-chunk-2}Pareto Exponent\label{zipf}}

@@

Pareto exp. & 2000 & 2001 & 2002 & 2003 & 2004 & 2005 & 2006 & 2007 & 2008 & 2009 & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018\\

@@

\begin{table}[!htbp] \centering
  \caption{Descriptive statistics for the multi-country model\label{desc.global}}
  \label{}
\footnotesize
\begin{tabular}{@{\extracolsep{1pt}}lccccc}
\\[-1.8ex]\hline
\hline \\[-1.8ex]
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Max} \\
\hline \\[-1.8ex]
Pareto exponent & 950 & $-$1.9 & $-$1.1 & 0.2 & $-$0.7 \\
Pareto inv. sq. error\textsuperscript{a} & 950 & 0.01 & 0.1 & 0.05 & 0.2 \\
Population density & 931 & 2.5 & 126.2 & 182.1 & 1,239.6 \\
Government expenditure (\% GDP) & 913 & 1.0 & 14.7 & 4.9 & 30.0 \\
Trade (\% GDP) & 914 & 19.1 & 65.3 & 33.9 & 220.4 \\
GDP growth & 934 & $-$33.1 & 4.0 & 4.2 & 54.2 \\
GDP per capita (log) & 892 & 630.7 & 18,049.2 & 15,432.9 & 61,391.4 \\
Female labour force (\%) & 950 & 7.9 & 37.8 & 11.9 & 55.2 \\
Population & 950 & 5,122,493 & 116,073,037.0 & 244,323,372.0 & 1,392,730,000 \\
Non agriculture value added (\% GDP) & 930 & 58.8 & 90.3 & 8.2 & 99.4 \\
Internet users per 100 hab. & 916 & 0.01 & 32.5 & 28.2 & 96.0 \\
Broadband users per 100 hab. & 824 & 0.0 & 8.8 & 11.2 & 44.8 \\
Mobile phone users per 100 hab. & 949 & 0.0 & 72.9 & 46.4 & 191.0 \\
Fixed phone users per 100 hab. & 946 & 0.0 & 19.8 & 18.6 & 68.4 \\
\hline \\[-1.8ex]
\multicolumn{6}{l}{\textsuperscript{a}This is the inverse squared standard error of the estimated Pareto exponent, which is used for weighting} \\
\multicolumn{6}{l}{the observations for the estimation, of equation 3. See Section 4} \\
\end{tabular}
\end{table}

@@

The availability of a panel dataset for city sizes across countries
enables us to use country fixed effects, which can address potential
endogeneity issues related to unobserved country specific
characteristics of city size distributions. However, such a strategy
does not address potential simultaneity issues. Simply put, internet
penetration might be affected by spatial structure, as reflected in the Pareto 
exponents, or both internet penetration and spatial structure might
be jointly determined by a third variable. E.g., if a country already
has a dispersed spatial structure, internet is particularly suitable in
facilitating communication. Potential endogeneity in our specification
will prevent us from being able to determine the causal impact of
internet and digital communication technologies usage on spatial
structure, which is the main aim of this paper. In order to address this
problem, we will adopt an instrumented variable strategy. Table
\ref{desc.global} also includes the descriptive statistics for the
instrumental variable we are using and (\(female\:labour\:force\)), will
be discussed in \protect\hyperlink{sec4}{Section 4}.

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the Pareto exponent, as estimated according to
the Gabaix and Ibragimov {[}44{]} correction. The main explanatory
variables of interest are provided at the top of Table \ref{ols.global};
namely internet, broadband, mobile and fixed telephony per \(100\)
habitants, expressed in natural logarithms. The main variables of
interest are introduced successively on their own in the regressions
reported in Table \ref{ols.global}. All regressions include country
fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared
standard error of the estimated  the Pareto exponent (see Table
\ref{desc.global}) to address potential noise that is carried over from
the first part of our identification strategy. In regards to the
interpretation of the estimated coefficients, given that the Pareto exponent
 has entered the regression not as an absolute value, but
instead as a real number a positive coefficient for a RHS variable
indicates an impact towards the decrease of the spatial dispersion of
population. In other words, a positive coefficient indicates an effect
towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies
because of the expansion of digital technologies.

\begin{table}[!htbp] \centering
  \caption{OLS estimation of equation (3)\label{ols.global}}
  \label{}
\small
\begin{tabular}{@{\extracolsep{1pt}}lcccc}
\\[-1.8ex]\hline
\hline \\[-1.8ex]
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\
\cline{2-5}
\\[-1.8ex] & \multicolumn{4}{c}{ Pareto Exponent 2000-18} \\
\\[-1.8ex] & (1) & (2) & (3) & (4)\\
\hline \\[-1.8ex]
 Internet users per 100 hab. (log) & 0.001$^{***}$ &  &  &  \\
  & (0.0001) &  &  &  \\
  Broadband users per 100 hab. (log) &  & $-$0.001$^{**}$ &  &  \\
  &  & (0.0002) &  &  \\
  Mobile phone users per 100 hab. (log) &  &  & 0.0003 &  \\
  &  &  & (0.0003) &  \\
  Fixed phone users per 100 hab. (log) &  &  &  & $-$0.00003 \\
  &  &  &  & (0.0001) \\
  Population density (log) & $-$1.378 & $-$1.060 & $-$1.505 & $-$1.892 \\
  & (1.870) & (1.985) & (1.896) & (2.007) \\
  Government expenditure (% GDP) & 0.0002 & 0.003$^{***}$ & 0.001 & 0.001 \\
  & (0.001) & (0.001) & (0.001) & (0.001) \\
  Trade (\% of GDP) & 0.0002$^{***}$ & $-$0.0001 & 0.0001 & 0.0001 \\
  & (0.0001) & (0.0001) & (0.0001) & (0.0001) \\
  Non agriculture value added (\% GDP) & 0.0001 & 0.001 & $-$0.0001 & $-$0.0002 \\
  & (0.001) & (0.001) & (0.001) & (0.001) \\
  GDP growth & 0.001$^{***}$ & 0.001$^{***}$ & 0.001$^{**}$ & 0.001$^{**}$ \\
  & (0.0003) & (0.0004) & (0.0004) & (0.0004) \\
  GDP per capita (log) & $-$0.020$^{***}$ & $-$0.023$^{***}$ & $-$0.019$^{***}$ & $-$0.016$^{**}$ \\
  & (0.007) & (0.006) & (0.007) & (0.008) \\
  Population (log) & 1.300 & 0.955 & 1.390 & 1.785 \\
  & (1.873) & (1.985) & (1.899) & (2.011) \\
  Constant & $-$18.795 & $-$13.969 & $-$19.948 & $-$25.516 \\
  & (26.314) & (27.886) & (26.670) & (28.261) \\
 \hline \\[-1.8ex]
Country fixed effects & Yes & Yes & Yes & Yes \\
Yearly fixed effects & Yes & Yes & Yes & Yes \\
Observations & 844 & 757 & 865 & 867 \\
Adjusted R$^{2}$ & 0.987 & 0.988 & 0.986 & 0.986 \\
Residual Std. Error & 0.516 & 0.497 & 0.540 & 0.540 \\
\hline
\hline \\[-1.8ex]
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\
 & \multicolumn{4}{r}{Robust Std. Errors in parenthesis} \\
\end{tabular}
\end{table}

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the Pareto exponent, might be affected by another source,
which also affects internet penetration. For instance, economic
development might affect the concentration of population in large cities
and at the same time enable more people to go online. If we do not
address this issue, the coefficient for the main variable of interest
will capture potential effects that internet penetration has on spatial
structure, but also potential reverse causality effects that spatial
structure might generate on internet penetration. To overcome this
potential problem, Table \ref{2sls.global} reports estimates of equation
\ref{eq3} using two-stage least squares (2SLS) with instrumental
variables (IVs). The latter are variables which are correlated with our
endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any
-- of the internet and digital communications on spatial structure. At a
first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs
and the other control variables are used instead of the endogenous
variable to estimate eq. \ref{eq3}. A significant effect will verify the
causal impact of the internet and digital communication usage on spatial
structure.

\begin{table}[!htbp] \centering
  \caption{2SLS estimation of equation (3)\label{2sls.global}}
  \label{}
\small
\begin{tabular}{@{\extracolsep{1pt}}lcccc}
\\[-1.8ex]\hline
\hline \\[-1.8ex]
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\
\cline{2-5}
\\[-1.8ex] & \multicolumn{4}{c}{ Pareto exponent, 2000-18} \\
\\[-1.8ex] & (1) & (2) & (3) & (4)\\
\hline \\[-1.8ex]
 Internet users per 100 hab. (log) & 0.001$^{***}$ &  &  &  \\
  & (0.0004) &  &  &  \\
  Broadband users per 100 hab. (log) &  & 0.002$^{**}$ &  &  \\
  &  & (0.001) &  &  \\
  Mobile phone users per 100 hab. (log) &  &  & $-$0.007$^{*}$ &  \\
  &  &  & (0.004) &  \\
  Fixed phone users per 100 hab. (log) &  &  &  & $-$0.001$^{***}$ \\
  &  &  &  & (0.0003) \\
  Population density (log) & $-$0.827 & $-$1.736 & $-$12.765 & $-$0.828 \\
  & (1.841) & (2.622) & (8.326) & (2.262) \\
  Government expenditure (\% GDP) & $-$0.0003 & 0.0002 & $-$0.002 & $-$0.0003 \\
  & (0.001) & (0.002) & (0.002) & (0.001) \\
  Trade (\% of GDP) & 0.0004$^{***}$ & 0.0004$^{*}$ & 0.001 & $-$0.0003 \\
  & (0.0001) & (0.0002) & (0.0005) & (0.0002) \\
  Non agriculture value added (\% GDP) & 0.0003 & $-$0.0003 & $-$0.001 & $-$0.001 \\
  & (0.001) & (0.001) & (0.001) & (0.001) \\
  GDP growth & 0.001$^{***}$ & 0.002$^{***}$ & 0.001$^{*}$ & 0.0001 \\
  & (0.0004) & (0.001) & (0.001) & (0.0004) \\
  GDP per capita (log) & $-$0.023$^{***}$ & 0.003 & 0.034 & $-$0.0003 \\
  & (0.007) & (0.012) & (0.038) & (0.011) \\
  Population (log) & 0.776 & 1.867 & 12.794 & 0.740 \\
  & (1.841) & (2.646) & (8.410) & (2.267) \\
  Constant & $-$11.513 & $-$27.468 & $-$180.823 & $-$10.915 \\
  & (25.852) & (37.246) & (118.577) & (31.852) \\
 \hline \\[-1.8ex]
Weak instruments & 44.77 & 22.23 & 3.51 & 27.24 \\
Wu-Hausman & 3 & 15.96 & 12.26 & 12 \\
P-value & 0.08 & 0 & 0 & 0 \\
Country fixed effects & Yes & Yes & Yes & Yes \\
Yearly fixed effects & Yes & Yes & Yes & Yes \\
Observations & 844 & 757 & 865 & 867 \\
Adjusted R$^{2}$ & 0.986 & 0.984 & 0.944 & 0.979 \\
Residual Std. Error & 0.534 & 0.579 & 1.078 & 0.657 \\
\hline
\hline \\[-1.8ex]
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\
 & \multicolumn{4}{r}{Robust Std. Errors in parenthesis} \\
 & \multicolumn{4}{r}{IV: Female participation in labour force} \\
\end{tabular}
\end{table}

@@

In total, the results indicate that our measures of internet and
telephony penetration have further enhanced agglomeration forces, at
least for large urban agglomerations, while the opposite happens for
fixed telephony. These results are robust against endogeneity issues,
but are limited to urban agglomeration included in our data. Indeed, for
the estimation of  the Pareto exponent we only included urban
agglomeration of \(300,000\) inhabitants or more due to data
availability. Hence, the above estimations cannot verify whether such an
effect is also valid for smaller cities. Interestingly, the above
results regarding the role of fixed telephony are in agreement with
previous results from {[}18{]} {[}19{]} which used a smaller panel of
countries but without constraints regarding the city size. However, our
results do not support the findings of {[}19{]} regarding internet
usage.

It is indeed puzzling that the results on Table \ref{2sls.global} that the effects on urban concentration for internet users and broadband users, on the one hand, and mobile and fixed telephony users are in opposite directions. We wish to broach this subject in the context of some related literature, namely Calabrese \emph{et al.} (2011)

Interplay between Telecommunications and Face-to-Face Interactions: A Study Using Mobile Phone Data

Francesco Calabrese ,Zbigniew Smoreda,Vincent D. Blondel,Carlo Ratti

Published: July 13, 2011https://doi.org/10.1371/journal.pone.0020814

and

B\"{u}chel and Ehrlich

Cities and the Structure of Social Interactions:
Evidence from Mobile Phone Data

http://staff.vwi.unibe.ch/vonehrlich/Papers/BuchelEhrlich012019.pdf

The latter supports the notion that ICT favors agglomeration in the sense that in spite of their convenience modern technologies do no make up for the need of proximity in  social interactions. The former supports the notion people who communicate more frequently are likely to be near one another. ICT, as viewed through such granular studies of individuals' communications does not make up for physical distance; individuals nearer one another communicate more.

In order to overcome the city size limitation, the next section
presents two case studies, for which we have obtained much more granular
data and therefore we are in a position to test the effect of the
internet on the tail of the urban population distribution for these
countries.

@@

We report estimation results using international country-level data on
urban agglomerations with more than \(300,000\) inhabitants from the UN
Urban Settlements data and test the effect of ICT adoption on the Pareto exponents for national city size distributions, as
measures of dispersion for heavy-tailed data. We report results which
are robust to endogeneity concerns. Then, in order to examine such
effects for cities that are smaller than those included in our
international data, we focus on the US and the UK urban systems.
Specifically, we test the effect of internet usage and internet speed on
the changes over time in rankings of Micropolitan and Metropolitan Areas
in the US, and of Built Up Areas in the UK. The results favor a
complementary relation between the internet and agglomeration
externalities. While the cross-country estimates indicate that increase
adoption rates of such technologies has resulted in less dispersed urban
spatial structures, the two case studies reveal that internet adoption
and internet speed improved the relative population rank of a city
within its urban system. Interestingly, the latter results indicate that
such effects might be even stronger for smaller and less dense urban
areas at least in the UK. On the contrary, our estimates regarding the
effect of fixed telephony advocate towards the reverse effect: that is
fixed telephony acting against of agglomeration forces.

@@

\leavevmode\hypertarget{ref-gabaix_rank_2011}{}%
44. Gabaix X, Ibragimov R. Rank- 1/2: A simple way to improve the OLS
estimation of tail exponents. Journal of Business \& Economic
Statistics. 2011;29: 24--39.
